## From a raw effect to a standardized effect size

::::: columns
::: {.column width="50%"}
`b_Experimental_condition` is on the scale of the writing score

<br>

To compare with the literature we want a [standardized effect size]{.hl}

<br>

`Cohen's d = b / SD`

<br>

Do this [for every draw]{.hl-blue} → a posterior distribution of *d*
:::

::: {.column width="50%"}
<b>But: which SD?</b>

-   `sigma`: residual SD, after pretest & classes

-   `sqrt(sd_Class__Intercept^2 + sigma^2)`: total SD in the model

-   pooled SD of `SecondVersion` (observed): the classic Cohen's *d*

::: {.calcnote}
The numerator is the [pretest-adjusted]{.hl} difference. The denominator is a choice, so always report which SD you used.
:::
:::
:::::

## Effect sizes based on the draws

```{r}
#| echo: true
#| message: false
#| code-line-numbers: "2|6|7|8"
#| fig-width: 10
#| fig-height: 4.6
#| out-width: "90%"
#| output-location: slide
# observed pooled within-condition SD of the posttest
SD_pooled <- sigma(lm(SecondVersion ~ factor(Experimental_condition), data = WritingData))

posterior_d <- posterior_PD %>%
  mutate(
    d_residual = b_Experimental_condition / sigma,
    d_total    = b_Experimental_condition / sqrt(sd_Class__Intercept^2 + sigma^2),
    d_observed = b_Experimental_condition / SD_pooled
  ) %>%
  select(d_residual, d_total, d_observed) %>%
  pivot_longer(everything(), names_to = "denominator", values_to = "d")

ggplot(posterior_d, aes(x = d, y = denominator)) +
  stat_halfeye(.width = c(.50, .89)) +
  geom_vline(xintercept = c(-.1, .1), linetype = "dashed") +  # ROPE
  scale_x_continuous("Cohen's d", breaks = seq(-.1, 1.2, by = .1)) +
  scale_y_discrete("", limits = c("d_observed", "d_total", "d_residual"))
```

## Effect sizes as probabilities

<br>

The draws let us answer: [how likely is the effect at least small / medium?]{.hl}

<br>

```{r}
#| echo: true
#| message: false
posterior_d %>%
  group_by(denominator) %>%
  summarise(
    median   = median(d),
    lower_89 = quantile(d, .055),
    upper_89 = quantile(d, .945),
    `P(d > 0.2)` = mean(d > .2),
    `P(d > 0.5)` = mean(d > .5)
  )
```
