---
title: "Applied Bayesian Analyses in R"
subtitle: "Appendices"
author: "Sven De Maeyer"
format: 
  revealjs:
    theme: [default, ../edubron-slides.scss, ../../_uantwerpen-brand.scss]
    width: 1422
    height: 805
    slide-number: true
self-contained: true
execute: 
  echo: false
  include: true
  output: true
code:
  code-copy: true
  code-line-numbers: true
code-annotations: hover
---

```{r}
#| echo: false
#| include: false
#| message: false
#| warning: false

library(tidyverse)

theme_set(theme_linedraw(base_size = 15))

NAVY <- "#002e65"
RED  <- "#ea2c38"
GREY <- "#5a6b80"

# ── One made-up study, used on every slide in this appendix ──────────
# A remedial programme is given to the pupils who started low. By the
# posttest they have caught up. Nothing exotic: a pretest, a posttest
# and a condition. Both tests are turned into z scores, as they would be
# in the example.
set.seed(2026)
n <- 60

d <- tibble(
  cond    = rep(c("control", "programme"), each = n),
  pre_raw = rnorm(2 * n, mean = ifelse(cond == "programme", -1, 1), sd = 0.6)
) |>
  mutate(
    post_raw = 0.9 * pre_raw + 1.6 * (cond == "programme") + rnorm(2 * n, sd = 0.40),
    pretest  = as.vector(scale(pre_raw)),
    posttest = as.vector(scale(post_raw)),
    cond     = factor(cond, levels = c("control", "programme"))
  )

m1 <- lm(posttest ~ pretest, data = d)
m2 <- lm(posttest ~ pretest + cond, data = d)

b_pooled   <- unname(coef(m1)["pretest"])
b_adjusted <- unname(coef(m2)["pretest"])
b_dummy    <- unname(coef(m2)["condprogramme"])

# how far apart are the two conditions, before and after?
gap <- function(v) abs(diff(tapply(v, d$cond, mean)))
gap_pre  <- unname(gap(d$pretest))
gap_post <- unname(gap(d$posttest))

# the pre-post relationship INSIDE the conditions
r_within <- cor(resid(lm(pretest ~ cond, d)), resid(lm(posttest ~ cond, d)))

# the same model, but on a properly randomised study
set.seed(7)
nr <- 60
r <- tibble(
  cond    = rep(c("control", "programme"), each = nr),
  pre_raw = rnorm(2 * nr)
) |>
  mutate(
    post_raw = 0.8 * pre_raw + 0.5 * (cond == "programme") + rnorm(2 * nr, sd = 0.55),
    pretest  = as.vector(scale(pre_raw)),
    posttest = as.vector(scale(post_raw)),
    cond     = factor(cond, levels = c("control", "programme"))
  )

rnd_pooled   <- unname(coef(lm(posttest ~ pretest, data = r))["pretest"])
rnd_adjusted <- unname(coef(lm(posttest ~ pretest + cond, data = r))["pretest"])
rnd_gap_pre  <- unname(abs(diff(tapply(r$pretest,  r$cond, mean))))
rnd_gap_post <- unname(abs(diff(tapply(r$posttest, r$cond, mean))))

XL <- c(-2.6, 2.6)
YL <- c(-3.0, 3.0)
```

# Can a standardized coefficient be larger than 1?

## The set-up

A pretest and a posttest, both turned into [z scores]{.hl}.

An experiment with [two conditions]{.hl}, so the model carries a dummy as well.

. . .

$$\text{posttest} = \text{pretest} + \text{condition}$$

. . .

The coefficient for the pretest. Is it stuck between [-1 and +1]{.hl}?

## First instinct: yes

```{r}
#| fig-height: 4.2
#| fig-width: 8.5
#| fig-align: center
ggplot(d, aes(pretest, posttest)) +
  geom_point(colour = GREY, alpha = .55, size = 2.6) +
  geom_abline(intercept = coef(m1)[1], slope = b_pooled,
              colour = NAVY, linewidth = 1.3) +
  coord_cartesian(xlim = XL, ylim = YL) +
  labs(x = "pretest (z)", y = "posttest (z)")
```

With [only the pretest]{.hl} in the model the coefficient is `r round(b_pooled, 2)`.

That number is literally the correlation between the two tests, so it can never pass 1.

# Now put the condition in the model {background-color="#002e65" transition="slide-in"}

## The same data, read a second way

```{r}
#| fig-height: 4.2
#| fig-width: 8.5
#| fig-align: center
ggplot(d, aes(pretest, posttest, colour = cond)) +
  geom_point(alpha = .6, size = 2.6) +
  geom_abline(intercept = coef(m2)[1], slope = b_adjusted,
              colour = NAVY, linewidth = 1.3) +
  geom_abline(intercept = coef(m2)[1] + b_dummy, slope = b_adjusted,
              colour = RED, linewidth = 1.3) +
  scale_colour_manual(values = c(control = NAVY, programme = RED)) +
  coord_cartesian(xlim = XL, ylim = YL) +
  labs(x = "pretest (z)", y = "posttest (z)", colour = NULL) +
  theme(legend.position = "top")
```

Same points. The coefficient went from `r round(b_pooled, 2)` to [`r round(b_adjusted, 2)`]{.hl}.

## Nothing odd happened to the data

Inside each condition, the pretest and the posttest correlate at [`r round(r_within, 2)`]{.hl}.

A perfectly ordinary number. Nothing above 1 anywhere in the data.

. . .

So the coefficient did not become large because the relationship became impossibly strong.

It became large because [the question it answers changed]{.hl}.

## Here is what actually changed

```{r}
#| fig-height: 4.0
#| fig-width: 10
#| fig-align: center
gaps <- d |>
  select(cond, pretest, posttest) |>
  pivot_longer(c(pretest, posttest), names_to = "moment", values_to = "z") |>
  mutate(moment = factor(moment,
                         levels = c("pretest", "posttest"),
                         labels = c("at the start", "at the end")))

gmeans <- gaps |>
  group_by(moment, cond) |>
  summarise(z = mean(z), .groups = "drop")

ggplot(gaps, aes(z, cond, colour = cond)) +
  geom_jitter(width = 0, height = .18, alpha = .40, size = 2) +
  geom_point(data = gmeans, size = 6, shape = 18) +
  scale_colour_manual(values = c(control = NAVY, programme = RED), guide = "none") +
  facet_wrap(~ moment, ncol = 1) +
  coord_cartesian(xlim = XL) +
  labs(x = "z score", y = NULL) +
  theme(panel.grid.major.y = element_blank())
```

The conditions start [`r round(gap_pre, 2)` standard deviations apart]{.hl}
and finish [`r round(gap_post, 2)` standard deviations apart]{.hl}.

## In one sentence

<br>

### The dummy ate [most of the pretest differences]{.hl} and [almost none of the posttest differences]{.hl}.

<br>

. . .

The coefficient is a slope, so it is a rise divided by a run.

Putting the condition in the model took away most of the [run]{.hl} and left the [rise]{.hl} alone.

## Why that lifts the number above 1

Standardizing makes the [whole group]{.hl} one standard deviation wide, on both tests.

That is what keeps the coefficient inside -1 and +1 when the pretest stands alone.

. . .

Once the condition is in the model, the coefficient only uses the variation [inside]{.hl} each condition.

But it is still reported in whole group units.

. . .

Two different rulers. [That is the whole trick.]{.hl}

## A study that produces this

A remedial programme is offered to the pupils who [started low]{.hl}.

By the posttest they have [caught up]{.hl} with everyone else.

. . .

Ignore the condition and the two tests look only mildly related, because the
group that scored low at the start no longer scores low at the end.

. . .

Put the condition in the model and the relationship inside each group becomes
visible again. The pooled picture was hiding it.

## In a randomised experiment this does not happen

Random assignment means the conditions [start in the same place]{.hl}.

| | without the dummy | with the dummy | gap at the start | gap at the end |
|---|---|---|---|---|
| remedial study | `r round(b_pooled, 2)` | [`r round(b_adjusted, 2)`]{.hl} | `r round(gap_pre, 2)` SD | `r round(gap_post, 2)` SD |
| randomised study | `r round(rnd_pooled, 2)` | `r round(rnd_adjusted, 2)` | `r round(rnd_gap_pre, 2)` SD | `r round(rnd_gap_post, 2)` SD |

. . .

In the randomised study the dummy barely touches the coefficient, because there
were no pretest differences for it to take away.

## So if your coefficient passes 1

Do not report a remarkably strong pretest effect.

. . .

Go and look at [the balance at the start]{.hl}.

The model is telling you that your conditions were [further apart on the pretest than they ended up on the posttest]{.hl}.

. . .

In a randomised design that is a question about the randomisation.

In a quasi experiment it is usually the design doing exactly what it was meant to do.

## And for your priors

The coefficient can in principle be any number, so a prior does not need a wall at 1.

A prior centred somewhere around 0.5 with a sensible width says what you believe
about pre and post designs, and still leaves room for a surprise.

::: {.calcnote}
**For the curious.** The adjusted coefficient equals the within condition
correlation, multiplied by the share of posttest spread that survives the dummy
and divided by the share of pretest spread that survives it. It passes 1 exactly
when the conditions differ more at the start than at the end. Everything on these
slides follows from that one ratio, and nothing else is needed to read it.
:::
